Generation of TypeScript Declaration Files from JavaScript Code

### 2021-05-17 mplr-2021/main.tex
Comments May 18

We aim to improve on this situation by providing a tool that generates
TypeScript declaration files automatically. To do so, we rely on the examples provided
by the developer as part of the documentation of a library. If these
examples are sufficiently rich, then our toolchain can generate
high-quality declaration files for the library. This way we take
advantage of 
best practices in the dynamic languages community which favors
examples and tests over writing out type signatures.

@@

Our tool \texttt{dts-generate} is a first step to explore the possibilities for
generating useful declaration files from code
examples. \texttt{dts-generate} comes with a framework that 
supports the generation of declaration files for an existing
JavaScript library published to the \NPM{} registry. The tool gathers
data flow and type information at run time to generate a declaration
file based on that information.

@@

Like \texttt{dts-gen}, a TypeScript package that generates a template for a declaration
file, we determine the kind of declaration file by examining the usage
of the module. If the
imported entity is used solely as a function as in Figure~\ref{fig:example-module-function}, the
generated declaration file follows the \textbf{module-function} template as shown in Figure~\ref{fig:template-module-function}.
If the example only accesses properties of the imported entity as in
Figure~\ref{fig:example-module}, we generate according to the \textbf{module} template (Figure~\ref{fig:template-module}).
If the imported entity is used with \lstinline/new/ to create new instances as in
Figure~\ref{fig:example-class}, we generate a \textbf{module-class} template (Figure~\ref{fig:template-class}).
While it is possible to create a library of undetermined category, our selection is driven
by the examples in the documentation and 
hence reflects the developers intent.
% Run-time analysis proves useful in this case,
% since it is not sufficient to statically analyze the exported module.

@@

@@

In the end, option~\ref{item:3} was the most viable option even though there
is no standard for documentation, either. However, almost all repositories
contain README files where the library authors briefly describe in prose what
the code does, which problem it solves, how to install the
application, how to build the code, etc. It is very common that
developers provide code examples in the README files to show how the
library works and how to use it. These code fragments showcase common use cases rather
than stress-testing the implementation (the problem of
option~\ref{item:2}) because they are meant to be instructive to users of the package.

@@

@@

The property \texttt{requiredModule} stores the name of the module that declared the invoked function. If a function is explicitly 
exported by the module, the property \texttt{isExported} will be set to \texttt{true} when it is invoked. If a function of an 
exported object is invoked, \texttt{isExported} will be \texttt{false} and \texttt{requiredModule} will contain the name of 
the required module.

@@

We were unable to  generate a declaration file based on the
\textbf{module-class} template. All modules corresponding to this template exhibited at least one of
the unimplemented TypeScript features.

>>>>

% I think the reader gets the impression from the text that obtaining the code examples
% is a "main feature" of our tool, where in fact it is only a way of easily obtaining
% runtime information to execute our tool. We could've followed the approach of obtaining
% the runtime information from the tests.
% This can be useful also to divert the attention from how many repositories had README files, etc.
% After using the code examples, we also found out that they provided a useful approximation
% not only on the input types but also on how to use the library (i.e. we don't analyze the structure of the library).
% This is a useful insight which can be used later to infer the correct types.
%
% Should we change this to 'we rely on runtime information'. 'We used the examples
% provided by the developer as part of the documentation of a library as a first
% pragmatic way of obtaining runtime information'.
We aim to improve on this situation by providing a tool that generates
TypeScript declaration files automatically. To do so, we rely on the examples provided
by the developer as part of the documentation of a library. If these
examples are sufficiently rich, then our toolchain can generate
high-quality declaration files for the library. This way we take
advantage of 
best practices in the dynamic languages community which favors
examples and tests over writing out type signatures.

@@

Our tool \texttt{dts-generate} is a first step to explore the possibilities for
generating useful declaration files from code
examples.
% Should we change it to 'explore the possibilities for
% generating useful declaration files from runtime information'.

\texttt{dts-generate} comes with a framework that 
supports the generation of declaration files for an existing
JavaScript library published to the \NPM{} registry. The tool gathers
data flow and type information at run time to generate a declaration
file based on that information.

@@

Unlike \texttt{dts-gen}, a TypeScript package that generates a template for a declaration
file which only analyzes the structure of the module, we determine the kind of declaration file by examining the usage
of the module. If the
imported entity is used solely as a function as in Figure~\ref{fig:example-module-function}, the
generated declaration file follows the \textbf{module-function} template as shown in Figure~\ref{fig:template-module-function}.
If the example only accesses properties of the imported entity as in
Figure~\ref{fig:example-module}, we generate according to the \textbf{module} template (Figure~\ref{fig:template-module}).
If the imported entity is used with \lstinline/new/ to create new instances as in
Figure~\ref{fig:example-class}, we generate a \textbf{module-class} template (Figure~\ref{fig:template-class}).
While it is possible to create a library of undetermined category, our selection is driven
by the examples in the documentation and 
hence reflects the developers intent.
% Run-time analysis proves useful in this case,
% since it is not sufficient to statically analyze the exported module.

@@

% I think we should rephrase this, specially now that we know
% it might be 'doable'.

@@

% We can maybe change the focus to 'Option 3' was easier to implement.
% This can be another opportunity to mention that  
In the end, option~\ref{item:3} was the most viable option to easily gather runtime information
even though there is no standard for documentation, either. However, almost all repositories
contain README files where the library authors briefly describe in prose what
the code does, which problem it solves, how to install the
application, how to build the code, etc. It is very common that
developers provide code examples in the README files to show how the
library works and how to use it. These code fragments showcase common use cases rather
than stress-testing the implementation (the problem of
option~\ref{item:2}) because they are meant to be instructive to users of the package.

@@

\begin{figure*}[t]
  \begin{subfigure}[t]{0.48\linewidth}
    \begin{lstlisting}[language=JavaScript,numbers=none]
const foo = require('foo');

foo.doSomething();
    \end{lstlisting}
  \caption{Example code for \texttt{module} template}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
    \begin{lstlisting}[language=JavaScript,numbers=none,escapechar=!]
{
  functionId_1: {
    functionName: 'doSomething',
    //...
    !\textbf{isExported: false,}!
    !\textbf{requiredModule: 'foo'}!
  }
}    
    \end{lstlisting}
    \caption{Runtime information fragment for \texttt{module} template}
  \end{subfigure}

  \begin{subfigure}[t]{0.48\linewidth}
    \begin{lstlisting}[language=JavaScript,numbers=none]
const foo = require('foo');

foo();
    \end{lstlisting}
    \caption{Example code for \texttt{module-function} template}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
    \begin{lstlisting}[language=JavaScript,numbers=none,escapechar=!]
{
  functionId_1: {
    functionName: 'foo',
    //...
    !\textbf{isExported: true,}!
    !\textbf{requiredModule: 'foo'}!
  }
}    
    \end{lstlisting}
    \caption{Runtime information for \texttt{module-function} template}
  \end{subfigure}
  \caption{Example code for properties \texttt{isExported} and \texttt{requiredModule} - Property \texttt{isExported} is
  \texttt{false} for \texttt{module} and \texttt{true} for \texttt{module-function}. Additionally, \texttt{requiredModule}
  is \texttt{foo} for both cases.
  }
  \label{fig:example-is-exported-required-module}
\end{figure*}

@@

The property \texttt{requiredModule} stores the name of the module that declared the invoked function. If a function is explicitly 
exported by the module, the property \texttt{isExported} will be set to \texttt{true} when it is invoked. If a function of an 
exported object is invoked, \texttt{isExported} will be \texttt{false} and \texttt{requiredModule} will contain the name of 
the required module, as shown in \figref{fig:example-is-exported-required-module}.

@@

% This is no longer true. I'll add one result to the report.
We were unable to  generate a declaration file based on the
\textbf{module-class} template. All modules corresponding to this template exhibited at least one of
the unimplemented TypeScript features.
